---
title: "Machine Learning Final"
subtitle: "Exposome Dataset, Option C"
author: "DC, AN, NP"
date: "4/25/2022"
always_allow_html: true
output:
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Research Question: Does prenatal exposure to alcohol influence birth weight outcomes among infants using matching by propensity scores?

**Option C **
Construct a propensity score to aid in an epidemiologic analysis of an etiologic question and then perform an analysis using the propensity score to address that etiologic question (choice of weighting or matching is yours)

**Relevance to Public Health:** 
This research question is important for public health because it adds to the existing body of literature regarding the relationship between prenatal alcohol exposure and health outcomes. There is inconsistent literature with respect to a causal relationship between maternal prenatal alcohol exposure and low birth weight in infants. In one study by Lundbergs et al., low to moderate prenatal alcohol consumption during early pregnancy and third trimester were associated with lower odds of low birth weight. A systematic review of RCTs and observational studies using alternative analytical approaches to improve causal inference by Mamluk et al. found some evidence of reduced birth weight with prenatal alcohol exposure. This research question is important for public health because it adds to the existing body of literature regarding the relationship between prenatal alcohol exposure and health outcomes. Since it would be unethical to randomize prenatal alcohol exposure, other methods must be used to improve causal inference. Given that other covariates may predict prenatal alcohol use, such as sociodemographic factors, utilization of a propensity score for prenatal alcohol exposure further elucidate the relationship between prenatal alcohol exposure and birth weight.

**RQ Type**
Explanatory 

**Citations:** 
Lundsberg LS, Illuzzi JL, Belanger K, Triche EW, Bracken MB. Low-to-moderate prenatal alcohol consumption and the risk of selected birth outcomes: a prospective cohort study. Ann Epidemiol. 2015;25(1):46-54.e3. doi:10.1016/j.annepidem.2014.10.011

Mamluk L, Jones T, Ijaz S, Edwards HB, Savović J, Leach V, Moore THM, von Hinke S, Lewis SJ, Donovan JL, Lawlor DA, Smith GD, Fraser A, Zuccolo L, Evidence of detrimental effects of prenatal alcohol exposure on offspring birthweight and neurodevelopment from a systematic review of quasi-experimental studies, Int J Epidemiol. 2020;49(6):1972-1995. doi.org/10.1093/ije/dyz272


### Data description 
Dataset: Exposome 
233 Variables, 1,301 Observations 

Potential propensity score features:
Outcome: 
e3_bw = birthweight(g)

Exposure: 
e3_alcpreg_yn_None = Alcohol during pregnancy (yes/no)

Covariates: 
h_age_None = Maternal age
hs_wgtgain_None = Maternal weight gain during pregnancy (kg)
h_folic_t1_None = folic acid supplementation during pregnancy
h_mbmi_None = Maternal pre-pregnancy body mass index (kg/m2)
e3_sex_None = Sex of child 
e3_gac_None = Gestational age


```{r packages, include=FALSE, results='hide'}
library(tidyverse)
library(modelr)
library(caret)
library(stats)
library(gtsummary)
library(broom)
library(randomForest)
library(MatchIt)
library(factoextra)
library(ggbiplot)
library(knitr)
```

#### Read in data set, merge, and clean 
```{r data prep, results='hide'}
#Load data 
load("./exposome.RData")

#Merge all data into single dataframe
studydata<-merge(exposome,phenotype,by="ID") %>% 
  merge(covariates, by="ID")

#Remove ID feature  
studydata$ID<-NULL

#Restrict to only needed variables
studydata_clean <- studydata %>%
  select(c("e3_bw", "e3_alcpreg_yn_None", "h_age_None", "hs_wgtgain_None", "h_folic_t1_None", "h_mbmi_None", "e3_sex_None", "e3_gac_None")) %>%
  mutate(birthweight_cat = if_else(e3_bw >= 2500, "normal birthweight", "low birthweight"), 
         birthweight_cat = as.factor(birthweight_cat))
#Convert to factor -> all applicable variables are in factor format, move to next step 

#Inspect data 
str(studydata_clean) %>% 
  knitr::kable()

```


## Construct a Propensity Score

### Propensity score using logit 
```{r prop logit, results='hide'}
ps.model.logit <- glm(e3_alcpreg_yn_None ~ h_age_None + hs_wgtgain_None + h_folic_t1_None + h_mbmi_None + e3_sex_None + e3_gac_None, 
                    data=studydata_clean, family=binomial(link="logit"))

#summary(ps.model.logit)
    
#Estimates odds of alcohol use, then convert to probability (aka the propensity score)
prop.score <- (predict(ps.model.logit, studydata_clean, type="response"))

#The logistic regression estimated PS
studydata_clean$PS.LOGIT <- prop.score 

```
```{r}
broom::glance(ps.model.logit) %>% 
  knitr::kable(digits = 3) %>% 
  kableExtra::kable_styling(full_width = F)

broom::tidy(ps.model.logit) %>% 
  select( -statistic) %>% 
  knitr::kable(digits = 3) %>% 
  kableExtra::kable_styling(full_width = F)
```

### Propensity score using random forest
```{r prop score rf, results='hide'}
set.seed(100)

feat.count<-c((ncol(studydata_clean)-4), (ncol(studydata_clean)-4)/2, sqrt(ncol(studydata_clean)-4))

grid.rf<-expand.grid(mtry=feat.count)

tree.num<-seq(100,500, by=200)

results.trees<-list()

for (ntree in tree.num){
 set.seed(100)
  rf.train<-train(e3_alcpreg_yn_None ~ h_age_None + hs_wgtgain_None + h_folic_t1_None + h_mbmi_None + e3_sex_None + e3_gac_None, 
                    data=studydata_clean, method="rf", metric="Accuracy", tuneGrid=grid.rf, importance=TRUE, ntree=ntree)
index<-toString(ntree)
results.trees[[index]]<-rf.train$results
}

output.trees<-bind_rows(results.trees, .id = "ntrees")
best.tune<-output.trees[which.max(output.trees[,"Accuracy"]),]

ps.model.rf<-randomForest(e3_alcpreg_yn_None ~ h_age_None + hs_wgtgain_None + h_folic_t1_None + h_mbmi_None + e3_sex_None + e3_gac_None, data=studydata_clean, mtry=3, ntree=500)

#Obtain propensity scores
ps.rf<-ps.model.rf$votes
studydata_clean$PS.RF<-ps.rf[,2]

#Compare propensity scores
plot(studydata_clean$PS.LOGIT, studydata_clean$PS.RF)
```

### Propensity score using elastic net
```{r prop score en, results='hide'}

lambda<-10^seq(-3,3, length=100)
alpha<-seq(0,1,by=0.1)

trnCtrl = trainControl(
             method = "CV",
             number = 10)

srchGrd = expand.grid(alpha=alpha, lambda=lambda)

ps.model.NET <- train(e3_alcpreg_yn_None ~ .,
                  data = studydata_clean,
                  method = "glmnet",
                  tuneGrid = srchGrd,
                  trControl = trnCtrl,
                  standardize = FALSE,
                  maxit = 1000000)

summary(ps.model.NET) %>% 
  knitr::kable()

prop.score.net <- predict(ps.model.NET, studydata_clean, type="prob") [,2]
studydata_clean$PS.NET <- prop.score.net

```

## Examine region of common support

```{r, message = FALSE}
ggplot(data=studydata_clean, aes(x=PS.LOGIT))+geom_histogram()+facet_grid(~e3_alcpreg_yn_None)+theme_bw()+ggtitle("Overlap PS from Logistic Regression")

ggplot(data=studydata_clean, aes(x=PS.RF))+geom_histogram()+facet_grid(~e3_alcpreg_yn_None)+theme_bw()+ggtitle("Overlap PS from Random Forest")

ggplot(data=studydata_clean, aes(x=PS.NET))+geom_histogram()+facet_grid(~e3_alcpreg_yn_None)+theme_bw()+ggtitle("Overlap PS from Elastic Net")
```


## Match by propensity score in 1:1 matching and compare covariate balance and population size

### LOGIT
```{r, results='hide'}
nn1 <- matchit(e3_alcpreg_yn_None ~ h_age_None + hs_wgtgain_None + h_folic_t1_None + h_mbmi_None + e3_sex_None + e3_gac_None, 
                    data=studydata_clean, distance=studydata_clean$PS.LOGIT, method="nearest", discard="both", caliper=0.2, ratio=1)

nn1.data <- match.data(nn1)
summary(nn1, standardize=T)
```

### RANDOM FOREST
```{r, results='hide'}
nn1.rf <- matchit(e3_alcpreg_yn_None ~ h_age_None + hs_wgtgain_None + h_folic_t1_None + h_mbmi_None + e3_sex_None + e3_gac_None, 
                    data=studydata_clean, distance=studydata_clean$PS.RF, method="nearest", discard = "both", caliper=0.2, ratio=1)
      
nn1.data.rf <- match.data(nn1.rf)
    
summary(nn1.rf, standardize=T)
```

### ELASTIC NET 
```{r, results='hide'}
nn1.net <- matchit(e3_alcpreg_yn_None ~ h_age_None + hs_wgtgain_None + h_folic_t1_None + h_mbmi_None + e3_sex_None + e3_gac_None, 
                    data=studydata_clean, distance=studydata_clean$PS.NET, method="nearest", discard = "both", caliper=0.2, ratio=1)
      
nn1.data.net <- match.data(nn1.net)
    
summary(nn1.net, standardize=T)
```

#### Average Standardized Mean Difference-Unmatched
```{r average SMD, results='hide'}

    # Original unmatched cohort
    ogasmd = mean(abs(summary(nn1, standardize=T)$sum.all[, 3][-1])) 
    
    # Matching attempt #1 Logistic Regression
    logitasmd = mean(abs(summary(nn1, standardize=T)$sum.matched[, 3][-1])) 

    # Matching attempt #2 Random Forest
    rfasmd = mean(abs(summary(nn1.rf, standardize=T)$sum.matched[, 3][-1])) 
    
    # Matching attempt #3 Elastic net
    enasmd = mean(abs(summary(nn1.net, standardize=T)$sum.matched[, 3][-1])) 
```

Model                      | Average SMD
-------------------------- | -------------
Original unmatched cohort  | 'r toString(ogasmnd)' 
Logistic Regression        | 'r toString(logitasmd)'
Random Forest              | 'r toString(rfasmd)'
Elastic Net                | 'r toString(enasmd)'


### Estimate effect for LOGIT because it had the lowest Avg SMD 
```{r}

nn1.data <- nn1.data %>%
  mutate(birthweight_cat = if_else(birthweight_cat == "normal birthweight", 0, 1))

outcome.model.1 <- glm(birthweight_cat ~ e3_alcpreg_yn_None, data=nn1.data, family=binomial(link="logit"))
    
  exp(outcome.model.1$coefficients) %>% 
    kable()
  exp(confint(outcome.model.1)) %>% 
    kable()
```


## Create new variable 
### Clean and tidy data
```{r, echo=FALSE, results='hide'}
#Strip PS vars from earlier, change all vars to numeric 
studydata_2 <- studydata_clean %>%
  select(-PS.LOGIT, -PS.RF, -PS.NET) %>%
  mutate(e3_sex_None = if_else(e3_sex_None == "male", 0, 1)) %>%
  mutate(hs_wgtgain_None = as.numeric(hs_wgtgain_None), 
         e3_alcpreg_yn_None = as.numeric(e3_alcpreg_yn_None), 
         h_folic_t1_None = as.numeric(h_folic_t1_None))

#Restrict to low bwt
studydata_restrict <-studydata_2[(which(studydata_clean$birthweight_cat=="low birthweight")),]

#Strip off the outcome from earlier 
studydata_features<-studydata_2 %>%
  select(-birthweight_cat)
```

### Use PCA to create new variable
```{r, echo=FALSE, results='hide'}

#Obtain and compare means and standard deviations across features. na.rm removes the missings
colMeans(studydata_features, na.rm = TRUE)
apply(studydata_features, 2, sd, na.rm=TRUE)

#Multiple differences in means, center and scale 
studydata.pca<-prcomp( ~., data=studydata_features, center=TRUE, scale=TRUE, na.action=na.omit)

#Can compare sds used to scale with the sds above to ensure they are close.
studydata.pca$scale

#Generates scree plot
fviz_eig(studydata.pca)

#view results of pca. Note the first three components are needed to explain at least 75% of the variance
summary(studydata.pca)

#Identify how features loaded on the different components
studydata.pca$rotation

ggbiplot(studydata.pca)

ggbiplot(studydata.pca, choices=c(2,3))

```









